Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
Minor change.
Section 9: refine.

Table~\ref{tab:defining-sustainability} reproduces a summary created
by C.~Lenhardt ~\cite{lenhardt-wssspe1-panel}, which shows for each
contribution to the Modeling Sustainability panel what it meant when
referring to software, how it defined software sustainability, and
which approach it suggested to understand or evaluate sustainability.
\begin{table}[t]
  \begin{scriptsize}
    \begin{center}
      \caption{Summary of Modeling Sustainability papers from Policy Panel.  Adapted from~\cite{lenhardt-wssspe1-panel}.}
      \label{tab:defining-sustainability}
      \begin{tabular}{|p{2.3cm}|p{3.6cm}|p{4.4cm}|p{4.8cm}|}
                \hline
{\bf Paper/Authors}
& {\bf Software}
& {\bf Sustainability}
& {\bf Approach to Understand or Evaluate Sustainability} \\
                \hline
Calero, et al.~\cite{Calero_WSSSPE}
& General notion of software. Not explicitly defined.
& Sustainability is linked to quality.
& Add to ISO \\
                \hline
Venters, et al.~\cite{Venters_WSSSPE}
& Software as science software; increasingly complex; service-oriented computing
& Extensibility, interoperability, maintainability, portability, reusability, scalability, efficiency
& Use various architecture evaluation approaches to assess sustainability \\
                \hline
Pierce, et al.~\cite{Pierce_WSSSPE}
& Cyberinfrastructure software
& Sustainable to the extent to which there is a community to support it
& Open community governance \\
                \hline
Katz and Proctor~\cite{Katz_WSSSPE}
& E-research infrastructures (i.e. cyberinfrastructure)
& Persisting over time, meeting original needs and projected needs
& Equates models for the creation of software with sustaining software \\
                \hline
Lenhardt,~et~al.~\cite{Lenhardt_WSSSPE}
& Broadly defined as software supporting science
& Re-use; reproducible science
& Comparing data management life cycle to software development life cycle \\
                \hline
Weber, et al.~\cite{Weber_WSSSPE}
& Software broadly defined; a software ecosystem
& Software niches
& Ecological analysis and ecosystem \\
                \hline
     \end{tabular}
    \end{center}
  \end{scriptsize}
\end{table}

@@

\subsubsection*{Use cases}
Two distinct class of scientific software projects and products were
discussed in the workshop: 1) generic, large-scale utility software
and 2) niche, medium- and small-scale scientific software. Each class
faces different and significant challenges. New undertakings should
recognize the differences in advance and identify such challenges
within the development and sustaining efforts. In particular, the
dynamics associated with developers, scope, life cycle,
users-community, (re)usability, funding support and career paths vary
widely among the two classes of software.

@@

\subsubsection*{Workshop process}
The WSSSPE workshop can be viewed as an experiment in how we can
collaboratively and inclusively build a workshop agenda, without
asking large numbers of people to submit papers that will be rejected
so that the workshop can have a low acceptance rate.

Contributors also want to get credit for their participation in the
process.  And the workshop organizers want to make sure that the
workshop content and their efforts are recorded.  The methods used in
the WSSSPE1 workshop were successful; we had good participation, many
contributors have a report they can cite, and the record of the
workshop is open and available, through the self-published reports,
the workshop web site and notes site, and this paper.  In addition,
many additional papers are being created that will include the
discussions at the workshop, including extended versions of many of
the self published reports. such as those that are in this special
issue.

@@

Additional activities that are being planned include two additional
WSSSPE workshops at the 2014 SciPy and SC14 conferences.  The SciPy
workshop (WSSSPE1.1) will focus on how some software projects from the
SciPy community have dealt with software sustainability issues, both
successfully and unsuccessfully, while the SC14 workshop (WSSSPE2)
will be more general, and will likely focus on determining and
publicizing specific activities that the overall scientific software
community can take to move forward.  In addition, there will be a
minisymposium on ``Reliable Computational Science'' at the 2014
Society for Industrial and Applied Mathematics Annual Meeting (SIAM
AN14, \url{http://meetings.siam.org}) to further explore some of the
key issues raised here.

>>>>

Table~\ref{tab:defining-sustainability} reproduces a summary created
by C.~Lenhardt ~\cite{lenhardt-wssspe1-panel}, which shows for each
contribution to the Modeling Sustainability panel what it meant when
referring to software, how it defined software sustainability, and
which approach it suggested to understand or evaluate sustainability.
\begin{table}[t]
  \begin{scriptsize}
    \begin{center}
      \caption{Summary of Modeling Sustainability papers from Policy Panel.  Adapted from~\cite{lenhardt-wssspe1-panel}.}
      \label{tab:defining-sustainability}
      \begin{tabular}{|p{2.3cm}|p{3.6cm}|p{4.4cm}|p{4.8cm}|}
                \hline
{\bf Paper/Authors}
& {\bf Software}
& {\bf Sustainability}
& {\bf Approach to Understand or Evaluate Sustainability} \\
                \hline
Calero, et al.~\cite{Calero_WSSSPE}
& General notion of software. Not explicitly defined.
& Sustainability is linked to quality.
& Add to ISO \\
                \hline
Venters, et al.~\cite{Venters_WSSSPE}
& Software as science software; increasingly complex; service-oriented computing
& Extensibility, interoperability, maintainability, portability, reusability, scalability, efficiency
& Use various architecture evaluation approaches to assess sustainability \\
                \hline
Pierce, et al.~\cite{Pierce_WSSSPE}
& Cyberinfrastructure software
& Sustainable to the extent to which there is a community to support it
& Open community governance \\
                \hline
Katz, et al.~\cite{Katz_WSSSPE}
& E-research infrastructures (i.e. cyberinfrastructure)
& Persisting over time, meeting original needs and projected needs
& Equates models for the creation of software with sustaining software \\
                \hline
Lenhardt,~et~al.~\cite{Lenhardt_WSSSPE}
& Broadly defined as software supporting science
& Re-use; reproducible science
& Comparing data management life cycle to software development life cycle \\
                \hline
Weber, et al.~\cite{Weber_WSSSPE}
& Software broadly defined; a software ecosystem
& Software niches
& Ecological analysis and ecosystem \\
                \hline
     \end{tabular}
    \end{center}
  \end{scriptsize}
\end{table}

@@

\subsubsection*{Use cases}
Two distinct class of scientific software projects and products were
discussed in the workshop: 1) generic, large-scale utility software
and 2) niche, medium- and small-scale scientific software. Each class
faces different and significant challenges. New undertakings should
recognize the differences in advance and identify such challenges
within the development and sustaining efforts. In particular, the
dynamics associated with developers, scope, life cycle,
users-community, (re)usability, funding support, and career paths vary
widely among the two classes of software.

@@

\subsubsection*{Workshop process}
The WSSSPE workshop can be viewed as an experiment in how we can
collaboratively and inclusively build a workshop agenda, without
asking a large number of people to submit papers that will be rejected
so that the workshop can have a low acceptance rate.

Contributors also want to get credit for their participation in the
process.  And the workshop organizers want to make sure that the
workshop content and their efforts are recorded.  The methods used in
the WSSSPE1 workshop were successful: we had good participation;  
contributors have a report they can cite;   the record of the
workshop is open and available through the self-published reports,
the workshop website and notes site, and this paper.  In addition,
many additional papers are being created that will include the
discussions at the workshop, including extended versions of many of
the self-published reports such as those that are in this special
issue.

@@

Additional activities that are being planned include two additional
WSSSPE workshops at the 2014 SciPy and SC14 conferences.  The SciPy
workshop (WSSSPE1.1) will focus on how some software projects from the
SciPy community have dealt with software sustainability issues, both
successfully and unsuccessfully, while the SC14 workshop (WSSSPE2)
will be more general, and will likely focus on determining and
publicizing specific activities that the overall scientific software
community can take to move forward.  In addition, there will be a
two-session minisymposium on ``Reliable Computational Science'' at the 2014
Society for Industrial and Applied Mathematics Annual Meeting (SIAM
AN14, \url{http://meetings.siam.org}) to further explore some of the
key issues raised here.
